Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces

### 2024-06-16 EBDVF-2024/paper.qmd
Add files via upload

---
title: Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces
author:
  - name: Robert David
    email: robert.david@semantic-web.com
    orcid: 0000-0002-3244-5341
    corresponding: true
    affiliations:
      - id: swc
        name: The Semantic Web Company
        country: Austria
  - name: Petar Ivanov
    email: petar.ivanov@ontotext.com
    orcid: 0000-0001-8448-1005
    affiliations:
      - id: onto
        name: Ontotext
        country: Bulgaria
  - name: Vladimir Alexiev
    email: vladimir.alexiev@ontotext.com
    orcid: 0000-0001-7508-7428
    corresponding: true
    affiliations:
      - ref: onto
abstract: |
  Dataspaces are an important enabler for industrial sharing data (either commercially licensed or private).
  Europe is investing heavily into sectoral dataspaces, federation and orchestration platforms like SIMPL, Eclipse DSC, GXFS, etc.
  Still, dataspaces enable shared data access, but do not solve the data interoperability problem.
  For that, the consumer would like to see the data from different providers in a harmonized and semantically integrated form.
  The Vocabulary Hub service (part of the IDS RAM) provides a repository for ontologies and vocabularies.
  We describe an approach of raising the role of the vocabulary hub to also allow richer metadata description (e.g. the meaning of every column in a tabular dataset),
  and binding semantic descriptions to ingested datasets, thus providing on-the-fly data semantization and easing data querying.
  This is achieved through the integration of two commercial semantic products (PoolParty and GraphDB),
  leveraging the partnership between the Semantic Web Company and Ontotext,
  and is being developed within the frame of the Digital Europe project UNDERPIN,
  with applications to refinery and wind farm data.
keywords:
  - dataspaces
  - semantic interoperability
  - semantic technologies
  - ontologies
  - vocabulary hub
  - oil and gas
  - renewable energy
  - refineries
  - windfarms
journal:
  highlights: # This makes a first page only with the Elsevier highlights: maybe we should remove it
    - The Vocabulary Hub is used in a dataspace to hold semantic vocabularies (ontologies and thesauri).
    - Proposes and implements a Semantic Layers architecture that allows binding semantic vocabularies to incoming datasets.
    - Uses semantic mapping languages (OBDA, R2RML, CSVW) to harmonize datasets dynamically.
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Workshop on Semantic Interoperability in Data Spaces
      formatting: preprint
      model: 3p
      cite-style: super
---

@@

Other architecture reference models, such as the GAIA-X Architecture [@GaiaXArchitectureDocument2022] 
describe components with a similar role: the Federated Catalogue and Broker Service.
Therefore our approach  is adaptable and applicable to dataspaces designed in accordance with other reference architecture models as well.

@@

## PoolParty Semantic Suite
[PoolParty Semantic Suite](PoolParty Semantic Suite) is a semantic middleware platform based on W3C Semantic Web standards 
that provides a wide variety of functionality in the area of knowledge graph management, graph-based NLP, semantic search and recommender systems.

@@

In the following, we discuss
i) the services provided for the Vocabulary Hub component and
ii) additional services as Vocabulary Hub extensions that implement a Semantic Layer for dataspaces.

@@

The Data Consumer benefits from:
- More powerful dataset discovery by using richer semantic queries, e.g.
  - "Give me all time series related to compressor C123" (equipment instance)
  - "Give me all time series related to temperature" (measured quantity)
  - "Give me all time series of bearings" (equipment kind or machine part)
- Harmonized querying of datasets from multiple providers.

>>>>

---
title: Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces
author:
  - name: Robert David
    email: robert.david@semantic-web.com
    orcid: 0000-0002-3244-5341
    corresponding: true
    affiliations:
      - id: swc
        name: The Semantic Web Company
        country: Austria
  - name: Petar Ivanov
    email: petar.ivanov@ontotext.com
    orcid: 0000-0001-8448-1005
    affiliations:
      - id: onto
        name: Ontotext
        country: Bulgaria
  - name: Vladimir Alexiev
    email: vladimir.alexiev@ontotext.com
    orcid: 0000-0001-7508-7428
    corresponding: true
    affiliations:
      - ref: onto
abstract: |
  Dataspaces are an important enabler for industrial sharing data (either commercially licensed or private).
  Europe is investing heavily into sectoral dataspaces, federation and orchestration platforms like SIMPL, Eclipse DSC, GXFS, etc.
  Still, dataspaces enable shared data access, but do not solve the data interoperability problem.
  For that, the consumer would like to see the data from different providers in a harmonized and semantically integrated form.
  The Vocabulary Hub service (part of the IDS RAM) provides a repository for ontologies and vocabularies.
  We describe an approach of raising the role of the Vocabulary Hub to also allow richer metadata description (e.g. the meaning of every column in a tabular dataset),
  and binding semantic descriptions to ingested datasets, thus providing on-the-fly data semantization and easing data querying.
  This is achieved through the integration of two commercial semantic products (PoolParty and GraphDB),
  leveraging the partnership between the Semantic Web Company and Ontotext,
  and is being developed within the frame of the Digital Europe project UNDERPIN,
  with applications to refinery and wind farm data.
keywords:
  - dataspaces
  - semantic interoperability
  - semantic technologies
  - ontologies
  - vocabulary hub
  - oil and gas
  - renewable energy
  - refineries
  - windfarms
journal:
  highlights: # This makes a first page only with the Elsevier highlights: maybe we should remove it
    - The Vocabulary Hub is used in a dataspace to hold semantic vocabularies (ontologies and thesauri).
    - Proposes and implements a Semantic Layers architecture that allows binding semantic vocabularies to incoming datasets.
    - Uses semantic mapping languages (OBDA, R2RML, CSVW) to harmonize datasets dynamically.
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Workshop on Semantic Interoperability in Data Spaces
      formatting: preprint
      model: 3p
      cite-style: super
---

@@

Other architecture reference models, such as the GAIA-X Architecture [@GaiaXArchitectureDocument2022] 
describe components with a similar role, namely the Federated Catalogue.
Therefore our approach is adaptable and applicable to dataspaces designed in accordance with other reference architecture models as well.

@@

## PoolParty Semantic Suite
[PoolParty Semantic Suite](https://www.poolparty.biz/) is a semantic middleware platform based on W3C Semantic Web standards 
that provides a wide variety of functionality in the area of knowledge graph management, graph-based NLP, semantic search and recommender systems.

@@

In the following, we discuss

i) the services provided for the Vocabulary Hub component and
ii) additional services as Vocabulary Hub extensions that implement a Semantic Layer for dataspaces.

@@

The Data Consumer benefits from:

- More powerful dataset discovery by using richer semantic queries, e.g.
  - "Give me all time series related to compressor C123" (equipment instance)
  - "Give me all time series related to temperature" (measured quantity)
  - "Give me all time series of bearings" (equipment kind or machine part)
- Harmonized querying of datasets from multiple providers.

### 2024-06-16 EBDVF-2024/paper.tex
Add files via upload

\begin{abstract}
Dataspaces are an important enabler for industrial sharing data (either
commercially licensed or private). Europe is investing heavily into
sectoral dataspaces, federation and orchestration platforms like SIMPL,
Eclipse DSC, GXFS, etc. Still, dataspaces enable shared data access, but
do not solve the data interoperability problem. For that, the consumer
would like to see the data from different providers in a harmonized and
semantically integrated form. The Vocabulary Hub service (part of the
IDS RAM) provides a repository for ontologies and vocabularies. We
describe an approach of raising the role of the vocabulary hub to also
allow richer metadata description (e.g.~the meaning of every column in a
tabular dataset), and binding semantic descriptions to ingested
datasets, thus providing on-the-fly data semantization and easing data
querying. This is achieved through the integration of two commercial
semantic products (PoolParty and GraphDB), leveraging the partnership
between the Semantic Web Company and Ontotext, and is being developed
within the frame of the Digital Europe project UNDERPIN, with
applications to refinery and wind farm data.
\end{abstract}

@@

Other architecture reference models, such as the GAIA-X Architecture
\citep{GaiaXArchitectureDocument2022} describe components with a similar
role: the Federated Catalogue and Broker Service. Therefore our approach
is adaptable and applicable to dataspaces designed in accordance with
other reference architecture models as well.

@@

\href{PoolParty\%20Semantic\%20Suite}{PoolParty Semantic Suite} is a
semantic middleware platform based on W3C Semantic Web standards that
provides a wide variety of functionality in the area of knowledge graph
management, graph-based NLP, semantic search and recommender systems.

@@

In the following, we discuss i) the services provided for the Vocabulary
Hub component and ii) additional services as Vocabulary Hub extensions
that implement a Semantic Layer for dataspaces.

@@

The Data Consumer benefits from: - More powerful dataset discovery by
using richer semantic queries, e.g. - ``Give me all time series related
to compressor C123'' (equipment instance) - ``Give me all time series
related to temperature'' (measured quantity) - ``Give me all time series
of bearings'' (equipment kind or machine part) - Harmonized querying of
datasets from multiple providers.

>>>>

\begin{abstract}
Dataspaces are an important enabler for industrial sharing data (either
commercially licensed or private). Europe is investing heavily into
sectoral dataspaces, federation and orchestration platforms like SIMPL,
Eclipse DSC, GXFS, etc. Still, dataspaces enable shared data access, but
do not solve the data interoperability problem. For that, the consumer
would like to see the data from different providers in a harmonized and
semantically integrated form. The Vocabulary Hub service (part of the
IDS RAM) provides a repository for ontologies and vocabularies. We
describe an approach of raising the role of the Vocabulary Hub to also
allow richer metadata description (e.g.~the meaning of every column in a
tabular dataset), and binding semantic descriptions to ingested
datasets, thus providing on-the-fly data semantization and easing data
querying. This is achieved through the integration of two commercial
semantic products (PoolParty and GraphDB), leveraging the partnership
between the Semantic Web Company and Ontotext, and is being developed
within the frame of the Digital Europe project UNDERPIN, with
applications to refinery and wind farm data.
\end{abstract}

@@

Other architecture reference models, such as the GAIA-X Architecture
\citep{GaiaXArchitectureDocument2022} describe components with a similar
role, namely the Federated Catalogue. Therefore our approach is
adaptable and applicable to dataspaces designed in accordance with other
reference architecture models as well.

@@

\href{https://www.poolparty.biz/}{PoolParty Semantic Suite} is a
semantic middleware platform based on W3C Semantic Web standards that
provides a wide variety of functionality in the area of knowledge graph
management, graph-based NLP, semantic search and recommender systems.

@@

In the following, we discuss

\begin{enumerate}
\def\labelenumi{\roman{enumi})}
\tightlist
\item
  the services provided for the Vocabulary Hub component and
\item
  additional services as Vocabulary Hub extensions that implement a
  Semantic Layer for dataspaces.
\end{enumerate}

@@

The Data Consumer benefits from:

\begin{itemize}
\tightlist
\item
  More powerful dataset discovery by using richer semantic queries, e.g.

  \begin{itemize}
  \tightlist
  \item
    ``Give me all time series related to compressor C123'' (equipment
    instance)
  \item
    ``Give me all time series related to temperature'' (measured
    quantity)
  \item
    ``Give me all time series of bearings'' (equipment kind or machine
    part)
  \end{itemize}
\item
  Harmonized querying of datasets from multiple providers.
\end{itemize}
